Deeply Integrating C11 Code Support into Isabelle/PIDE

### 2019-10-22 C11-FrontEnd/document/generated/paper.tex
synchronize with fortesse

  \textbf{In case that you consider citing Isabelle/C, please refer to \cite{Tuong-IsabelleC:2019}.  } %
\end{isamarkuptext*}\isamarkuptrue%
%
\begin{isamarkupsection*}%
[label = {intro},type = {scholarly_paper.introduction}, args={label = {intro},type = {scholarly_paper.introduction}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}}]Introduction%
\end{isamarkupsection*}\isamarkuptrue%
%
\begin{isamarkuptext*}%
[label = {introtext},type = {scholarly_paper.introduction}, args={label = {introtext},type = {scholarly_paper.introduction}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}}]\noindent{}Recent successes like the Microsoft Hypervisor project \cite{DBLP:conf/fm/LeinenbachS09},
the verified CompCert compiler \cite{DBLP:journals/cacm/Leroy09}
and the seL4 microkernel \cite{DBLP:conf/sosp/KleinEHACDEEKNSTW09,DBLP:journals/tocs/KleinAEMSKH14} 
show that the verification of low-level systems code has become feasible.
However, a closer look at the underlying verification engines  
VCC \cite{DBLP:conf/tphol/CohenDHLMSST09}, 
or Isabelle/AutoCorres \cite{DBLP:conf/pldi/GreenawayLAK14}
show that the road is still bumpy: the  empirical cost evaluation  of the L4.verified project 
\cite{DBLP:journals/tocs/KleinAEMSKH14} reveals that a very substantial part  of the overall 
effort of about one third of the 28 man years went into the development of libraries and the 
associated tool-chain. Accordingly, the project authors \cite{DBLP:journals/tocs/KleinAEMSKH14} 
express the hope that these overall investments will not have to be repeated for 
``similar projects''.

@@

[label = {hyplinkout},type = {Isa_COL.figure}, args={label = {hyplinkout},type = {Isa_COL.figure}, Isa_COL.figure.relative_width = {40}, Isa_COL.figure.src = {figures/markup-demo}, Isa_COL.figure.spawn_columns = {True}}]Output window showing the evaluation result of \autoref{fig:ml-pos} (clicking on the hyperlink $\protect\house$ makes the cursor jump to the \isatt{@{\char`\{}here{\char`\}}} of \autoref{fig:ml-pos})%

@@

[label = {lexer_ex},type = {scholarly_paper.technical}, args={label = {lexer_ex},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Preprocessing Lexical Conventions: Comments and Backslash Newlines%

@@

  $K_\text{\emph{cmd}}$ as : $K_\text{\emph{cmd}} = \isa{\isacommand{definition}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}defn} \isatt{|} \isa{\isacommand{lemma}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}goal{\isacharunderscore}stmt}

@@

  \caption{Activating the Isabelle/C/Clean back-end triggers the generation of theorems}

@@

  \caption{Isabelle/C/AutoCorres: integrating an HOL correctness proof near its C code}

@@

\item and the generated $\text{G}_{\text{ML}}$ depends on the two respective ASTs of Happy and
ML-Yacc.%
\end{itemize}%
As we assume the Isabelle code generator maintained by the Isabelle developers, and a
substantial part of Citadelle already maintained in the Isabelle AFP, the discussion remains on the
other tools used for generating $\text{AST}^{\text{C11}}$ and $\text{G}_{\text{ML}}$. For the case
of Happy and ML-Yacc, their own grammar languages have not significantly evolved in the last years,
and are still maintained in \url{https://github.com/simonmar/happy} and
\url{https://github.com/MLton/mlton}. For the case of Haskabelle, it has been
dropped from the Isabelle distribution, because it is critically relying on
haskell-src-exts~\url{http://hackage.haskell.org/package/haskell-src-exts}, and the
AST of haskell-src-exts received a drastic replacement from version 1.17.1 to 1.18.0.\footnote{See \url{http://hackage.haskell.org/package/haskell-src-exts-1.17.1} and
  \url{http://hackage.haskell.org/package/haskell-src-exts-1.18.0}. Actually, the
  default AST of haskell-src-exts was duplicated and enhanced in 2009 between 1.1.1 to 1.1.3 for a
  parsed source to carry additional position information in all its parsing nodes. Ultimately, the
  two ASTs was merged together while releasing version 1.18.0.} Consequently, upgrading
Haskabelle to use the new 1.18.0 represents an entire challenge. Hopefully, it is a task we already
completed while integrating Haskabelle's output with Citadelle's input. Otherwise, we have not
encountered any problems in upgrading Haskabelle from 1.18.0 to more recent versions (such as 1.20.1
which was straightforward).%
\isamarkupsubsection{A Gentle Zoom on Citadelle's Generator%
In more details, combining Haskabelle and Citadelle is almost direct, as Haskabelle's
output AST and Citadelle's input AST are already both subsets models of a rich Isar/HOL interface,
defined in past works~\cite{DBLP:journals/afp/TuongW15}. Note that one can configure Citadelle
for it to generate $\text{AST}^\text{C11}_{\text{HOL}}$ using
\isa{\isacommand{old{\isacharunderscore}datatype}} commands instead of
\isa{\isacommand{datatype}}, in particular to gain 10~min on the shallow-reflected
evaluation side (with the option \isa{quick{\isacharminus}and{\isacharminus}dirty} activated). In a nutshell, our new
implemented \isa{\isacommand{Haskell{\isacharunderscore}file}} command accepts other various
options:~\footnote{Note that \isa{\isacommand{old{\isacharunderscore}datatype}} is a command,
  whereas \isa{\isakeyword{datatype{\isacharunderscore}old}} is an option provided to the command
  \isa{\isacommand{Haskell{\isacharunderscore}file}}.}
\begin{tabular}{ll}
\isa{\isacommand{Haskell{\isacharunderscore}file}} & \isa{\isakeyword{datatype{\isacharunderscore}old}\ \isakeyword{try{\isacharunderscore}import}\ \isakeyword{only{\isacharunderscore}types}\ \isakeyword{concat{\isacharunderscore}modules}} \\
                & \isa{\isakeyword{base{\isacharunderscore}path}\ {\isachardoublequoteopen}{\isachardollar}HASKABELLE{\isacharunderscore}HOME{\isacharslash}ex{\isacharslash}language{\isacharminus}c{\isacharslash}src{\isachardoublequoteclose}} \\
                & \isa{{\isacharbrackleft}Prelude\ {\isasymrightharpoonup}\ C{\isacharunderscore}Model{\isacharunderscore}init{\isacharcomma}\ Int{\isacharcomma}\ String{\isacharcomma}\ Option\ {\isasymrightharpoonup}\ C{\isacharunderscore}Model{\isacharunderscore}init{\isacharbrackright}} \\
                & \isa{} \\
                & \isa{{\isachardoublequoteopen}{\isachardollar}HASKABELLE{\isacharunderscore}HOME{\isacharslash}ex{\isacharslash}language{\isacharminus}c{\isacharslash}src{\isacharslash}Language{\isacharslash}C{\isacharslash}Syntax{\isacharslash}AST{\isachardot}hs{\isachardoublequoteclose}}
\end{gather*}

The advantage of this construction is the possibility to do immediate code value transformations
obtained from Haskabelle: operationally, \isa{\isacommand{Haskell{\isacharunderscore}file}} makes an
implicit call to Haskabelle, translates the Haskell's output of Haskabelle to an acceptable HOL
input, and lets the Citadelle framework execute generated theorems and proofs in native concurrent
speed, based on a version of Isabelle specially modified for the purpose.\footnote{The use
  of a modified Isabelle is just optional: the whole Citadelle framework can also be configured to
  run with the standard version of Isabelle, i.e., at run-time the difference only relies on the
  performance treatment of consecutive commands.} On the one hand, the measure performance
of the \isa{\isacommand{datatype}} package is not improving when being generated and
executed by our Citadelle infrastructure, because the blocking expensive point remains located on

@@

command (which is not interacting with other commands). On the other hand, the question
of: \begin{quote}\emph{``How fast that specific \isa{\isacommand{datatype}}
  command would perform if from the 20 mutually recursive sub-definitions it were instead generating
  20 or several (sub-)commands to be executed in native concurrent speed?''}\end{quote} is
  actually opening an interesting topic of future experiments.%

@@

[label = {forest},type = {scholarly_paper.technical}, args={label = {forest},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Shift-Reduce forest of a parsing execution%

>>>>

  \textbf{In case that you consider to cite this work, please refer to \cite{Tuong-IsabelleC:2019}.}%
\end{isamarkuptext*}\isamarkuptrue%
%
\begin{isamarkupsection*}%
[label = {intro},type = {scholarly_paper.introduction}, args={label = {intro},type = {scholarly_paper.introduction}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}}]Introduction%
\end{isamarkupsection*}\isamarkuptrue%
%
\begin{isamarkuptext*}%
[label = {introtext},type = {scholarly_paper.introduction}, args={label = {introtext},type = {scholarly_paper.introduction}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}}]\noindent{}Recent successes like the Microsoft Hypervisor project \cite{DBLP:conf/fm/LeinenbachS09},
the verified CompCert compiler \cite{DBLP:journals/cacm/Leroy09}
and the seL4 microkernel \cite{DBLP:conf/sosp/KleinEHACDEEKNSTW09,DBLP:journals/tocs/KleinAEMSKH14} 
show that the verification of low-level systems code has become feasible.
However, a closer look at the underlying verification engines  
VCC \cite{DBLP:conf/tphol/CohenDHLMSST09}, 
or Isabelle/AutoCorres \cite{DBLP:conf/pldi/GreenawayLAK14}
show that the road is still bumpy: the  empirical cost evaluation  of the L4.verified project 
\cite{DBLP:journals/tocs/KleinAEMSKH14} reveals that a very substantial part  of the overall 
effort of about one third of the 28 man years went into the development of libraries and the 
associated tool-chain. Accordingly, the project authors \cite{DBLP:journals/tocs/KleinAEMSKH14} 
express the hope that these overall investments will not have to be repeated for 
``similar projects''.

@@

[label = {hyplinkout},type = {Isa_COL.figure}, args={label = {hyplinkout},type = {Isa_COL.figure}, Isa_COL.figure.relative_width = {40}, Isa_COL.figure.src = {figures/markup-demo}, Isa_COL.figure.spawn_columns = {True}}]Output window showing the evaluation result of \autoref{fig:ml-pos} (the anchor of this hyperlink
 $\protect\house$ targets at the position marked by \isatt{@{\char`\{}here{\char`\}}})%

@@

[label = {lexer_ex},type = {scholarly_paper.technical}, args={label = {lexer_ex},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Preprocessing Lexical Conventions: Comments and Newlines%

@@

  $K_\text{\emph{cmd}}$ as: $K_\text{\emph{cmd}} = \isa{\isacommand{definition}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}defn} \isatt{|} \isa{\isacommand{lemma}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}goal{\isacharunderscore}stmt}

@@

  \caption{Activating the Isabelle/C/Clean back-end \\ 
           triggers the generation of theorems}

@@

  \caption{Isabelle/C/AutoCorres: integrating an HOL correctness \\
           proof as annotation \emph{inside} C code}

@@

\item and the generated $\text{G}_{\text{ML}}$ depends on the two respective ASTs of Happy and
ML-Yacc.%
\end{itemize}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
As we assume the Isabelle code generator maintained by the Isabelle developers, and a
substantial part of Citadelle already maintained in the Isabelle AFP, the discussion remains on the
other tools used for generating $\text{AST}^{\text{C11}}$ and $\text{G}_{\text{ML}}$. For the case
of Happy and ML-Yacc, their own grammar languages have not significantly evolved in the last years,
and are still maintained in \url{https://github.com/simonmar/happy} and
\url{https://github.com/MLton/mlton}. For the case of Haskabelle, it has been
dropped from the Isabelle distribution, because it is critically relying on
haskell-src-exts~\url{http://hackage.haskell.org/package/haskell-src-exts}, and the
AST of haskell-src-exts received a drastic replacement from version 1.17.1 to 1.18.0.\footnote{See \url{http://hackage.haskell.org/package/haskell-src-exts-1.17.1} and
  \url{http://hackage.haskell.org/package/haskell-src-exts-1.18.0}. Actually, the
  default AST of haskell-src-exts was duplicated and enhanced in 2009 between 1.1.1 to 1.1.3 for a
  parsed source to carry additional position information in all its parsing nodes. Ultimately, the
  two ASTs was merged together while releasing version 1.18.0.} Consequently, upgrading
Haskabelle to use the new 1.18.0 represents an entire challenge. Hopefully, it is a task we already
completed while integrating Haskabelle's output with Citadelle's input. Otherwise, we have not
encountered any problems in upgrading Haskabelle from 1.18.0 to more recent versions (such as 1.20.1
which was straightforward).%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsection{A Zoom on Citadelle's HOL Generator%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
In the generation process of the parsers for Isabelle/C, the Citadelle 
system~\cite{DBLP:journals/afp/TuongW15} plays a vital role. In particular, it is used to 
convert Haskell AST presentations into HOL AST presentations through an initial call to 
Haskabelle. However, the resulting output can be run natively in order to increase efficiency
via Citadelle.

In more detail, combining Haskabelle and Citadelle is almost direct, as Haskabelle's
output AST and Citadelle's input AST are already both subset models of a rich Isar/HOL interface,
defined in past works~\cite{DBLP:journals/afp/TuongW15}. Note that one can configure Citadelle
to generate $\text{AST}^\text{C11}_{\text{HOL}}$ using \isa{\isacommand{old{\isacharunderscore}datatype}} commands instead of 
\isa{\isacommand{datatype}}. This results in a gain of 10~min on the shallow-reflected evaluation with the option 
\isa{quick{\isacharminus}and{\isacharminus}dirty} activated.

In the following, we show how we proceed: 
we implemented a command \isa{\isacommand{Haskell{\isacharunderscore}file}} that accepts  various
options:~\footnote{Note that \isa{\isacommand{old{\isacharunderscore}datatype}} is a command,
  whereas \isa{\isakeyword{datatype{\isacharunderscore}old}} is an option provided to the command
  \isa{\isacommand{Haskell{\isacharunderscore}file}}.}
\begin{gather*}
\begin{tabular}{ll}
\isa{\isacommand{Haskell{\isacharunderscore}file}} & \isa{\isakeyword{datatype{\isacharunderscore}old}\ \isakeyword{try{\isacharunderscore}import}\ \isakeyword{only{\isacharunderscore}types}\ \isakeyword{concat{\isacharunderscore}modules}} \\
                & \isa{\isakeyword{base{\isacharunderscore}path}\ {\isachardoublequoteopen}{\isachardollar}HASKABELLE{\isacharunderscore}HOME{\isacharslash}ex{\isacharslash}language{\isacharminus}c{\isacharslash}src{\isachardoublequoteclose}} \\
                & \isa{{\isacharbrackleft}Prelude\ {\isasymrightharpoonup}\ C{\isacharunderscore}Model{\isacharunderscore}init{\isacharcomma}\ Int{\isacharcomma}\ String{\isacharcomma}\ Option\ {\isasymrightharpoonup}\ C{\isacharunderscore}Model{\isacharunderscore}init{\isacharbrackright}} \\
                & \isa{} \\
                & \isa{{\isachardoublequoteopen}{\isachardollar}HASKABELLE{\isacharunderscore}HOME{\isacharslash}ex{\isacharslash}language{\isacharminus}c{\isacharslash}src{\isacharslash}Language{\isacharslash}C{\isacharslash}Syntax{\isacharslash}AST{\isachardot}hs{\isachardoublequoteclose}}
\end{tabular}
\end{gather*}

This construction has the advantage to allow immediate code value transformations
obtained from Haskabelle: operationally, \isa{\isacommand{Haskell{\isacharunderscore}file}} makes an
implicit call to Haskabelle, translates the output of Haskabelle to HOL
input, and lets the Citadelle framework execute the generated theorems and proofs in parallel.
For the latter, we use an Isabelle version that has been especially modified for this purpose.
\footnote{The use of a modified Isabelle is just an optimization: the whole Citadelle framework can 
also be configured to run on a standard Isabelle version, albeit with a significantly higher
execution time.}

 On the one hand, the execution of the AST mostly consisting of  \isa{\isacommand{datatype}} commands
 is not improving when being generated and
executed by our Citadelle infrastructure, because the blocking point is located in just

@@

command (which is not interacting with other commands). 
On the other hand, the question of advanced parallelisation techniques in the  \isa{\isacommand{datatype}}
package is actually opening an interesting topic of future experiments. 
The Citadelle infrastructure offers a number of mechanisms to address this task.%

@@

[label = {forest},type = {scholarly_paper.technical}, args={label = {forest},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Shift-Reduce Forest of a Parsing Execution%
